---
title: "Gambanalytics: Your guide to NFL Betting using NFL Trends and Data"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
---

# About

This book uses data from nflfastR (https://www.nflfastr.com/) and the R package. It is all a work in progress as well. 

Current plans:

- Add more visualization to team/QB Data

- Analyze season long stats, last 5 games and predict future games based on opponent

## Full Prediction Plans

Game Props:

- Game Result Money Line Prediction

- Spread Results

- Total Score Prediction

- Winning Margins

- Spread by Quarter

Player Props:

- Touchdown Scorer Prediction (First TD Scorer Potentially)

- Passing Yards O/U

- Pass Completions

- Pass Attempts

- Interception

- Passing TDs

- Longest Pass

- Rushing Yards O/U

- Longest Rush

- Rushing + Receiving Yards O/U

- Receiving Yards O/U

- Total Receptions

- Longest Reception

<!--chapter:end:index.Rmd-->

```{r include=FALSE}
library(nflfastR)
library(dplyr, warn.conflicts = FALSE)
library(tidyverse)
library(nflplotR)
source("buildFunctions.R")
knitr::opts_chunk$set(echo = TRUE)
nflreadr::.clear_cache()

offense_pbp <- load_pbp(2022) %>%
  filter(season_type == "REG") %>%
  filter(!is.na(posteam) & (rush == 1 | pass == 1))

scores <-
  offense_pbp[c(
    'game_id',
    'week',
    'home_team',
    'away_team',
    'season_type',
    'home_score',
    'away_score'
  )] %>%
  unique()

home_df <-
  scores[c('week', 'home_team', 'away_team', 'home_score', 'away_score')] %>%
  rename(team = home_team) %>%
  group_by(team) %>%
  mutate(
    oppositionTeam = away_team,
    PointsFor = mean(home_score),
    PointsAgainst = mean(away_score),
    win = if_else(home_score > away_score, 1, 0),
    pythWin = (PointsFor^13.91)/((PointsFor^13.91)+(PointsAgainst^13.91))
  ) %>%
  summarise(week,
            team,
            oppositionTeam,
            PointsFor,
            PointsAgainst,
            win,
            pythWin
            ) %>%
  unique()

away_df <-
  scores[c('week', 'away_team', 'home_team', 'home_score', 'away_score')] %>%
  rename(team = away_team) %>%
  group_by(team) %>%
  mutate(
    oppositionTeam = home_team,
    PointsFor = away_score,
    PointsAgainst = home_score,
    win = if_else(away_score > home_score, 1, 0),
    pythWin = (PointsFor^13.91)/((PointsFor^13.91)+(PointsAgainst^13.91))
  ) %>%
  summarize(week,
            team,
            oppositionTeam,
            PointsFor,
            PointsAgainst,
            win,
            pythWin
            ) %>%
  unique()

schedule_df <- rbind(x = away_df, y = home_df)
avg_points = mean(schedule_df$PointsFor)
```

# Team Data

The following are visuals and data that will help you with any Game Props or provide useful information to help you make more informed decisions when it comes to overall team-based results

## General Team Data

```{r warning = FALSE}
team_df <- schedule_df %>%
  group_by(team) %>%
  mutate(
    oRatio = mean(PointsFor / avg_points),
    dRatio = mean(PointsAgainst / avg_points),
    pointsFor = mean(PointsFor),
    pointsAgainst = mean(PointsAgainst),
    netPF = mean(PointsFor / dRatio),
    netPA = mean(PointsAgainst / oRatio),
    wins = sum(win),
    losses = sum(win == 0),
    winPct = wins / (wins + losses),
    pythWin = ((pointsFor^2.37)/((pointsFor^2.37)+(pointsAgainst^2.37)))*17
  ) %>%
  summarize(team,
            pointsFor,
            pointsAgainst,
            netPF,
            netPA,
            wins,
            pythWin,
            losses,
            winPct,
            oRatio,
            dRatio) %>%
  unique()

ggplot(team_df, aes(x = pointsFor, y = pointsAgainst)) +
  geom_mean_lines(aes(h_var = pointsFor, v_var = pointsAgainst)) +
  geom_nfl_logos(aes(team_abbr = team), width = 0.07, alpha = 0.7) +
  labs(x = "PointsFor/Game",
       y = "PointsAgainst/Game",
       title = "2021-2022 NFL Average Points For vs Points Against") +
  theme_bw() +
  theme(plot.title = element_text(size = 12, hjust = 0.5, face = "bold")) +
  scale_y_reverse()
```

```{r}
ggplot(team_df, aes(x = pointsFor, y = netPF)) +
  geom_mean_lines(aes(h_var = pointsFor, v_var = netPF)) +
  geom_nfl_logos(aes(team_abbr = team), width = 0.07, alpha = 0.7) +
  labs(x = "Actual Points For",
       y = "Weighted Points For",
       title = "2022-2023 NFL Points For vs Weighted Points For Based On Opponent") +
  theme_bw() +
  theme(plot.title = element_text(size = 12, hjust = 0.5, face = "bold"))

```

``` {r}
ggplot(team_df, aes(x = pointsAgainst, y = netPA)) +
  geom_mean_lines(aes(h_var = pointsAgainst, v_var = netPA)) +
  geom_nfl_logos(aes(team_abbr = team), width = 0.07, alpha = 0.7) +
  labs(x = "Actual Points Against",
       y = "Weighted Points Against",
       title = "2022-2023 NFL Points Against vs Weighted Points Against Based On Opponent") +
  theme_bw() +
  theme(plot.title = element_text(size = 12, hjust = 0.5, face = "bold"))+
  scale_y_reverse()+
  scale_x_reverse()
```

## Game Props Predictions




<!--chapter:end:01-intro.Rmd-->

```{r include=FALSE}
library(nflfastR)
library(dplyr, warn.conflicts = FALSE)
library(tidyverse)
library(nflplotR)
library(ggrepel)
library(knitr)
library(forcats)
library(ggpubr)
knitr::opts_chunk$set(echo = TRUE)
nflreadr::.clear_cache()
passing_pbp <- load_pbp(2022) %>%
  filter(season_type == "REG") %>%
  filter(!is.na(posteam) & (pass == 1) & (!is.na(passer)) & (!is.na(air_epa)) & (!is.na(xyac_epa)))
```
# QB Data

Here we will have all NFL QB-related Data and predictions

## General Data

```{r echo=FALSE, warning = FALSE}
epaPerPass <- passing_pbp %>%
  group_by(passer) %>%
  mutate(totalAttempts = n(),
         avg_ypa = mean(air_yards),
         avg_epa = mean(qb_epa),
         air_epa = mean(air_epa),
         yac_epa = mean(yac_epa),
         xyac_epa = mean(xyac_epa),
         xyac_yds = mean(xyac_mean_yardage),
         shortPassAtt = mean(air_yards < 10, na.rm = T),
         mediumPassAtt = mean((10 <= air_yards & air_yards <= 25), na.rm = T),
         deepPassAtt = mean(air_yards > 25, na.rm = T)
         ) %>%
  summarize(passer,
            totalAttempts,
            avg_ypa,
            avg_epa,
            air_epa,
            yac_epa,
            xyac_epa,
            xyac_yds,
            shortPassAtt,
            mediumPassAtt,
            deepPassAtt
            ) %>%
  filter(totalAttempts > 110) %>%
  unique()

shortAccuracy <- passing_pbp %>%
  filter(air_yards < 10) %>%
  group_by(passer) %>%
  mutate(
    shortCompPct = mean(complete_pass)
  ) %>%
  summarize(
    passer,
    shortCompPct
  ) %>%
  unique()

mediumAccuracy <- passing_pbp %>%
  filter(10 <= air_yards & air_yards <= 25) %>%
  group_by(passer) %>%
  mutate(
    mediumCompPct = mean(complete_pass)
  ) %>%
  summarize(
    passer,
    mediumCompPct
  ) %>%
  unique()

deepAccuracy <- passing_pbp %>%
  filter(air_yards > 25) %>%
  group_by(passer) %>%
  mutate(
    deepCompPct = mean(complete_pass)
  ) %>%
  summarize(
    passer,
    deepCompPct
  ) %>%
  unique()

epaPerPass <- epaPerPass %>%
  inner_join(shortAccuracy, by="passer") %>%
  inner_join(mediumAccuracy, by="passer") %>%
  inner_join(deepAccuracy, by="passer")

pbp <- passing_pbp %>%
  filter(passer %in% epaPerPass$passer)

ggplot(epaPerPass, aes(x = totalAttempts, y = avg_epa, label = passer)) +
  geom_jitter() +
  geom_label_repel(
    min.segment.length = 0,
    max.overlaps = Inf,
    label.size = 0,
    label.padding = 0,
    label.r = 0,
    size = 4
  ) +
  geom_smooth(
    method = lm,
    se = FALSE,
    col = 'red',
    linetype = 'dashed'
  )
```

```{r echo=FALSE, warning = FALSE}
epaPerPass <- passing_pbp %>%
  group_by(passer) %>%
  mutate(totalAttempts = n(),
         avg_ypa = mean(air_yards),
         avg_epa = mean(qb_epa),
         air_epa = mean(air_epa),
         yac_epa = mean(yac_epa),
         xyac_epa = mean(xyac_epa),
         xyac_yds = mean(xyac_mean_yardage),
         shortPassAtt = mean(air_yards < 10, na.rm = T),
         mediumPassAtt = mean((10 <= air_yards & air_yards <= 25), na.rm = T),
         deepPassAtt = mean(air_yards > 25, na.rm = T)
         ) %>%
  summarize(passer,
            totalAttempts,
            avg_ypa,
            avg_epa,
            air_epa,
            yac_epa,
            xyac_epa,
            xyac_yds,
            shortPassAtt,
            mediumPassAtt,
            deepPassAtt
            ) %>%
  filter(totalAttempts > 110) %>%
  unique()

shortAccuracy <- passing_pbp %>%
  filter(air_yards < 10) %>%
  group_by(passer) %>%
  mutate(
    shortCompPct = mean(complete_pass)
  ) %>%
  summarize(
    passer,
    shortCompPct
  ) %>%
  unique()

mediumAccuracy <- passing_pbp %>%
  filter(10 <= air_yards & air_yards <= 25) %>%
  group_by(passer) %>%
  mutate(
    mediumCompPct = mean(complete_pass)
  ) %>%
  summarize(
    passer,
    mediumCompPct
  ) %>%
  unique()

deepAccuracy <- passing_pbp %>%
  filter(air_yards > 25) %>%
  group_by(passer) %>%
  mutate(
    deepCompPct = mean(complete_pass)
  ) %>%
  summarize(
    passer,
    deepCompPct
  ) %>%
  unique()

epaPerPass <- epaPerPass %>%
  inner_join(shortAccuracy, by="passer") %>%
  inner_join(mediumAccuracy, by="passer") %>%
  inner_join(deepAccuracy, by="passer")

pbp <- passing_pbp %>%
  filter(passer %in% epaPerPass$passer)

ggplot(epaPerPass, aes(x = totalAttempts, y = avg_epa, label = passer)) +
  geom_jitter() +
  geom_label_repel(
    min.segment.length = 0,
    max.overlaps = Inf,
    label.size = 0,
    label.padding = 0,
    label.r = 0,
    size = 4
  ) +
  geom_smooth(
    method = lm,
    se = FALSE,
    col = 'red',
    linetype = 'dashed'
  )
```

```{r fig.width = 25,echo=FALSE, warning = FALSE}
data1 <- epaPerPass
long_sorted <- data1 # Replicate original data
long_sorted$passer <- factor(long_sorted$passer, # Factor levels in decreasing order
                  levels = long_sorted$passer[order(long_sorted$deepPassAtt, decreasing = TRUE)])

medium_sorted <- data1 
medium_sorted$passer <- factor(medium_sorted$passer,                                  
                 levels = medium_sorted$passer[order(medium_sorted$mediumPassAtt, decreasing = TRUE)])

short_sorted <- data1 
short_sorted$passer <- factor(short_sorted$passer,                                  
                 levels = short_sorted$passer[order(short_sorted$shortPassAtt, decreasing = TRUE)])

longPct_graph <- long_sorted %>%
  ggplot(aes(x = passer, y = deepPassAtt)) +
  ggtitle("Percentage of Pass Attempts >25 Yards") +
  geom_bar(stat = "identity", width = 0.7, fill = "#bc4b51") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

mediumPct_graph <- medium_sorted %>%
  ggplot(aes(x = passer, y = mediumPassAtt)) +
  ggtitle("Percentage of Pass Attempts within 10-25 Yards") +
  geom_bar(stat = "identity", width = 0.7, fill = "#f4a259") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

shortPct_graph <-short_sorted %>%
  ggplot(aes(x = passer, y = shortPassAtt)) +
  ggtitle("Percentage of Pass Attempts <10 Yards") +
  geom_bar(stat = "identity", width = 0.7, fill = "#8cb369") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

ggarrange(longPct_graph, mediumPct_graph, shortPct_graph,
          ncol = 1, nrow = 3)

```

```{r echo=FALSE, warning = FALSE}
ggplot(epaPerPass, aes(x = air_epa, y = yac_epa, label = passer)) +
  geom_jitter() +
  geom_label_repel(
    min.segment.length = 0,
    max.overlaps = Inf,
    label.size = 0,
    label.padding = 0,
    label.r = 0,
    size = 4
  ) +
  geom_smooth(
    method = lm,
    se = FALSE,
    col = 'red',
    linetype = 'dashed'
  )
```

```{r echo=FALSE, warning = FALSE}
ggplot(epaPerPass, aes(x = air_epa, y = avg_ypa, label = passer)) +
  geom_jitter() +
  geom_label_repel(
    min.segment.length = 0,
    max.overlaps = Inf,
    label.size = 0,
    label.padding = 0,
    label.r = 0,
    size = 4
  ) +
  geom_smooth(
    method = lm,
    se = FALSE,
    col = 'red',
    linetype = 'dashed'
  )
```

```{r fig.width = 25,echo=FALSE, warning = FALSE}
data1 <- epaPerPass
longComp_sorted <- data1 # Replicate original data
longComp_sorted$passer <- factor(longComp_sorted$passer, # Factor levels in decreasing order
                  levels = longComp_sorted$passer[order(longComp_sorted$deepCompPct, decreasing = TRUE)])

mediumComp_sorted <- data1 
mediumComp_sorted$passer <- factor(mediumComp_sorted$passer,                                  
                 levels = mediumComp_sorted$passer[order(mediumComp_sorted$mediumCompPct, decreasing = TRUE)])

shortComp_sorted <- data1 
shortComp_sorted$passer <- factor(shortComp_sorted$passer,                                  
                 levels = shortComp_sorted$passer[order(shortComp_sorted$shortCompPct, decreasing = TRUE)])

longCompPct_graph <- longComp_sorted %>%
  ggplot(aes(x = passer, y = deepCompPct)) +
  ggtitle("Completion% forf Pass Attempts >25 Yards") +
  geom_bar(stat = "identity", width = 0.7, fill = "#bc4b51") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

mediumCompPct_graph <- mediumComp_sorted %>%
  ggplot(aes(x = passer, y = mediumCompPct)) +
  ggtitle("Completion% for Pass Attempts within 10-25 Yards") +
  geom_bar(stat = "identity", width = 0.7, fill = "#f4a259") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

shortCompPct_graph <-shortComp_sorted %>%
  ggplot(aes(x = passer, y = shortCompPct)) +
  ggtitle("Completion% for Pass Attempts <10 Yards") +
  geom_bar(stat = "identity", width = 0.7, fill = "#8cb369") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

ggarrange(longCompPct_graph, mediumCompPct_graph, shortCompPct_graph,
          ncol = 1, nrow = 3)

```

```{r fig.weight=250,echo=FALSE, warning = FALSE}
shortPassScatter <- ggplot(epaPerPass, aes(x = shortPassAtt, y = shortCompPct, label = passer))+
    geom_jitter() +
  geom_label_repel(
    min.segment.length = 0,
    max.overlaps = Inf,
    label.size = 0,
    label.padding = 0,
    label.r = 0,
    size = 4
  ) +
  geom_smooth(
    method = lm,
    se = FALSE,
    col = 'red',
    linetype = 'dashed'
  )

mediumPassScatter <- ggplot(epaPerPass, aes(x = mediumPassAtt, y = mediumCompPct, label = passer))+
    geom_jitter() +
  geom_label_repel(
    min.segment.length = 0,
    max.overlaps = Inf,
    label.size = 0,
    label.padding = 0,
    label.r = 0,
    size = 4
  ) +
  geom_smooth(
    method = lm,
    se = FALSE,
    col = 'red',
    linetype = 'dashed'
  )

deepPassScatter <- ggplot(epaPerPass, aes(x = deepPassAtt, y = deepCompPct, label = passer))+
    geom_jitter() +
  geom_label_repel(
    min.segment.length = 0,
    max.overlaps = Inf,
    label.size = 0,
    label.padding = 0,
    label.r = 0,
    size = 4
  ) +
  geom_smooth(
    method = lm,
    se = FALSE,
    col = 'red',
    linetype = 'dashed'
  )
print(deepPassScatter)
print(mediumPassScatter)
print(shortPassScatter)

```

<!--chapter:end:02-cross-refs.Rmd-->

